Spatial Reasoning and Data Displays

### 2015-03-29 Paper/InfoVis2015/TestingVisualAptitude.Rnw
idea for the teaser

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
    \includegraphics[width=.4\linewidth]{lineup}
    \caption[Sample Lineup]{``Which of these plots is the most different from the others?"\newline
    This is the question participants are asked to answer for each of the lineups in the study. In the figure, a sample lineup of boxplots is shown. %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}}
  }

@@

In this paper we investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol  \cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

@@

\begin{figure}[ht]
\includegraphics[width=.9\linewidth]{Figure/fig-ResultsScaledScores-1}
\caption{Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \label{fig:Scores}}
\end{figure}
% 
\afterpage{\clearpage}

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=9,fig.height=5.5, message=F, warning=F, fig.pos='ht'>>=
loadfonts()
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "STEM Major", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}
\afterpage{\clearpage}

@@

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}
loadfonts()
require(grid)
ggplot()+
  geom_point(data=subset(all, points), aes(x=x, y=y), shape=1) + 
  geom_smooth(data=subset(all, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=subset(all, !points), aes(x=x, y=y, label=paste0("Correlation = \n", label))) + 
  geom_line(data=densities, aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank(),
        plot.margin=unit(c(0,0,0,0), unit="mm"))
rm(list=c("data", "all", "densities"))
embedFonts("./Figure/fig-VisReasoningSPM-1.pdf")
@
\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}
\afterpage{\clearpage}

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\afterpage{\clearpage}
%
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}

@@

\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-LineupTypeSPM-1}
\caption{Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \label{fig:lineuptaskmatrix}}
\end{figure}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure. Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks.

>>>>

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
\begin{minipage}[c]{.5\textwidth}
\includegraphics[width=\textwidth]{lineup}
\end{minipage}
\begin{minipage}[c]{.3\textwidth}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-1}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-2}
\end{minipage}
    \caption[Sample Lineup]{``Which of these plots is the most different from the others?"\newline
    This is the question participants are asked to answer for each of the lineups in the study. In the figure, a sample lineup of boxplots is shown. %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}
    The plots on the right show biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.}
  }

@@

Here, we want to investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol  \cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{Figure/fig-ResultsScaledScores-1}
\caption{Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \label{fig:Scores}}
\end{figure}
% 
%\afterpage{\clearpage}

@@

<<VisReasoningCategorical, echo=F, include=F, eval=T, fig.width=8.75,fig.height=6, message=F, warning=F, fig.pos='ht'>>=
loadfonts()
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "STEM Major", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="blank") +
  geom_boxplot(fill="grey", outlier.size=1.5) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=2.5) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}
%\afterpage{\clearpage}

@@

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}
loadfonts()
require(grid)
ggplot()+
  geom_point(data=subset(all, points), aes(x=x, y=y), shape=1) + 
  geom_smooth(data=subset(all, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=subset(all, !points), aes(x=x, y=y, label=paste0("Correlation = \n", label))) + 
  geom_line(data=densities, aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank(),
        plot.margin=unit(c(0,0,0,0), unit="mm"))
rm(list=c("data", "all", "densities"))
embedFonts("./Figure/fig-VisReasoningSPM-1.pdf")
@
\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}
%\afterpage{\clearpage}

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
%\afterpage{\clearpage}
%
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{Figure/fig-LineupTypeSPM-1}
\caption{Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \label{fig:lineuptaskmatrix}}
\end{figure}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure. Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks.
